---
title: "SVCFit Guide"
format: html
vignette: >
  %\VignetteIndexEntry{SVCFit Guide}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

SVCFit estimates the **Structural Variant Cellular Fraction (SVCF)** for somatic structural variants (SVs) detected in tumor sequencing data. SVCF is the proportion of all sampled cells that carry a given SV. Cancer cell fraction (CCF) instead uses cancer cells as its denominator and is obtained through the separately documented purity adjustment.

A key challenge is that SVs frequently co-occur with somatic copy-number alterations (CNAs). When an SV breakpoint falls inside a deleted or amplified segment, naively dividing SV-supporting reads by total reads gives a biased estimate of the true cellular fraction. SVCFit corrects for this by:

1.  Detecting whether a deletion or duplication overlaps each SV breakpoint.
2.  Using nearby heterozygous SNPs to infer the **zygosity** (hemizygous vs. homozygous) and **phase** (same vs. opposite allele as the CNA) of the SV.
3.  Applying the appropriate correction formula for each configuration.

The resulting `final_svcf` column is a copy-number–corrected cellular fraction that can be fed into downstream clustering and phylogenetic tree inference.

## Pipeline overview

```         
BAM (tumor + normal)
       │
       ├─▶ Manta → SV calls (VCF)
       │
       ├─▶ SVtyper → genotyped SV VCF (p_sv)
       │
       ├─▶ GATK HaplotypeCaller → het SNPs (p_het, p_onsv)
       │
       └─▶ FACETS → CNV segments (p_cnv)
                │
                ▼
          run_svcfit()
                │
                ▼
        SVCF table (.bed)
                │
                ▼
          build_trees()   ← optional: clustering + phylogenetic tree
```

------------------------------------------------------------------------

# Input files

SVCFit requires four input files. Example files ship with the package and are used throughout this vignette.

```{r}
#| label: setup
#| message: false
library(SVCFit)
library(dplyr)
library(tidyr)
library(ggplot2)
library(stringr)
library(readr)
library(GenomicRanges)

p_het   <- system.file("extdata", "examples/het_near_sv_c50p80m50.vcf",  package = "SVCFit")
p_onsv  <- system.file("extdata", "examples/het_on_sv_c50p80m50.vcf",   package = "SVCFit")
p_sv    <- system.file("extdata", "examples/example_sv.bed",             package = "SVCFit")
p_cnv   <- system.file("extdata", "examples/c50p80m50.bed",              package = "SVCFit")
p_truth <- system.file("extdata", "examples/",                           package = "SVCFit")
```

## `p_sv` — Structural variant calls

A headerless VCF produced by running **Manta** followed by **SVtyper** for genotyping. Each row is one SV breakpoint record with 11 tab-separated columns (10 for tumor-only mode).

```{r}
#| label: show-sv
#| echo: false
sv_raw <- read.table(p_sv, quote = '"')
colnames(sv_raw) <- c("CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO","FORMAT","normal","tumor")
head(sv_raw[, 1:8], 3)
```

Key columns:

| Column | Description |
|----------------------------|--------------------------------------------|
| `CHROM`, `POS` | Breakpoint chromosome and position |
| `ID` | Manta SV identifier (encodes SV type, e.g. `MantaDEL`, `MantaDUP`, `MantaINV`, `MantaBND`) |
| `ALT` | `<DEL>`, `<DUP:TANDEM>`, `<INV>`, or a breakend string for translocations |
| `QUAL` | SV quality score (low-quality calls are filtered by `QUAL_thresh`) |
| `INFO` | Contains `SVTYPE`, `SVLEN`, `END`, `SOMATIC`, etc. |
| `FORMAT` / `normal` / `tumor` | Genotype fields; SVtyper adds `PR` (paired reads) and `SR` (split reads) |

## `p_het` — Heterozygous SNPs near SVs

A headerless VCF from **GATK HaplotypeCaller** containing germline heterozygous SNPs in the tumor sample, restricted to a window around each SV breakpoint. Used to infer the zygosity and allelic phase of each SV.

```{r}
#| label: show-het
#| echo: false
het_raw <- read.table(p_het, quote = '"')
colnames(het_raw) <- c("CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO","FORMAT","bulk")
head(het_raw[, 1:7], 3)
```

## `p_onsv` — Heterozygous SNPs on SV-supporting reads

Same format as `p_het`. These are the same germline SNPs, but here only reads that support the SV (i.e. split reads or discordant read pairs crossing the breakpoint) are included. The ratio of the two allele counts on SV-supporting reads reveals which haplotype carries the SV.

## `p_cnv` — Copy-number segments (FACETS format)

A tab-separated file from **FACETS** (or any tool that produces the same columns). Each row is one segment.

```{r}
#| label: show-cnv
#| echo: false
cnv_raw <- read.delim(p_cnv, header = TRUE)
head(cnv_raw[, c("chrom","start","end","tcn.em","lcn.em","cf.em")], 4)
```

Key columns:

| Column | Description |
|----------------------------|--------------------------------------------|
| `chrom`, `start`, `end` | Segment coordinates |
| `tcn.em` | Total copy number of the segment |
| `lcn.em` | Minor-allele copy number (`NA` is treated as 0 by the current implementation) |
| `cf.em` | FACETS cellular fraction of the segment; retained as `cncf` for annotation and audit, but not used directly by `calc_svcf()` |

------------------------------------------------------------------------

# Quick start — `run_svcfit()`

For most users, a single call to `run_svcfit()` runs the full SVCF inference pipeline and returns a list with two elements: `svcf` (the SVCF result table) and `sv_info` (parsed SV metadata).

```{r}
#| label: quickstart
#| message: true
result <- run_svcfit(
  p_het   = p_het,
  p_onsv  = p_onsv,
  p_sv    = p_sv,
  p_cnv   = p_cnv,
  samp    = "example_sample",
  exper   = "exp1"
)

svcfit_out <- result$svcf
sv_info    <- result$sv_info
```

## Key output columns

```{r}
#| label: show-output
svcfit_out %>%
  select(ID, classification, bkg_cnv, cn_type, zygosity, final_svcf) %>%
  head(10)
```

| Column | Description |
|----------------------------|--------------------------------------------|
| `ID` | SV identifier from the input VCF |
| `classification` | SV type after processing: `DEL`, `DUP`, `INV`, `BND` (translocation), `COPBND` (copy-paste translocation) |
| `bkg_cnv` | Copy-number context at the breakpoint: `norm` (diploid), `del`, or `dup` |
| `cn_type` | Type of overlapping CNV event (`DEL`, `DUP`, or `norm`) |
| `zygosity` | Inferred allelic state of the SV: `het` (one allele) or `hom` (both alleles) |
| `final_svcf` | **Copy-number–corrected SVCF.** Ordinary valid estimates lie in 0–1; inspect `svcf_status` and `ss2_constraint_status` for exceptional and boundary cases. |
| `raw_svcf` | Initial class-adjusted SVCF estimate before an overlapping-CNA correction |
| `ss2_raw` | Unconstrained alternate overlapping-CNA SVCF candidate after the zygosity correction |
| `ss2_constraint_status` | Whether `ss2_raw` was in range or constrained to the lower or upper SVCF boundary |

### Interpreting `final_svcf`

This example data contains a artificial dataset with two clones (trunk 80% SVCF and subclone 40% SVCF).

```{r}
#| label: plot-svcf
#| fig-width: 5
#| fig-height: 3
ggplot(svcfit_out, aes(x = final_svcf, fill = classification)) +
  geom_histogram(binwidth = 0.05, colour = "white") +
  geom_density(
    aes(x = final_svcf, y = after_stat(density) * nrow(svcfit_out) * 0.05),
    inherit.aes = FALSE,
    colour = "black", linewidth = 0.8,
    adjust = 0.8
  ) +
  labs(x = "SVCF", y = "Count", title = "Distribution of SVCF estimates",
       fill = "SV type") +
  theme_bw()
```

------------------------------------------------------------------------

# Step-by-step walkthrough

This section breaks down what `run_svcfit()` does internally. Each sub-function can be called independently if you need finer control.

## Step 1: Extract information — `extract_info()`

`extract_info()` loads all four input files and parses them into structured data frames. It also processes translocations (BND records) and filters low-quality SVs.

```{r}
#| label: extract
#| message: false
info <- extract_info(
  p_het       = p_het,
  p_onsv      = p_onsv,
  p_sv        = p_sv,
  p_cnv       = p_cnv,
  chr_lst     = NULL,      # NULL = all chromosomes
  flank_del   = 50,        # bp window for BND-associated deletions
  QUAL_thresh = 100,       # minimum SV quality score
  min_alt     = 2,         # minimum SV-supporting reads
  tum_only    = FALSE      # paired normal BAM available
)

data      <- info[[1]]   # list(het_snp, het_on_sv, sv, cnv)
sv_info   <- info[[2]]   # parsed SV records
snp_df    <- info[[3]]   # parsed SNP records
sv_phase  <- info[[4]]   # SNP phase on SV reads
```

The input data has `r nrow(read.table(p_sv, quote='"'))` SV breakpoint records. After quality filtering, `r nrow(sv_info)` remain.

## Step 2: Characterize SVs — `characterize_sv()`

`characterize_sv()` integrates CNV segments with nearby heterozygous SNPs to annotate each SV with:

-   **Zygosity**: is the SV on one or both homologous chromosomes?
-   **Phase**: is the SV on the same haplotype as the overlapping CNA, or the opposite?
-   **Overlapping CNV**: which segment (if any) overlaps this SV breakpoint?

```{r}
#| label: characterize
anno_sv_cnv <- characterize_sv(sv_phase, sv_info, data$cnv)
head(anno_sv_cnv %>% select(ID, zygosity, ASCN, cn_type, bkg_cnv))
```

`ASCN` is the legacy column name for the SNP-derived **allele copy ratio (ACR)** relevant to the SV allele. It is a read-ratio estimate, not a FACETS integer allele-specific copy number.

## Step 3: Calculate SVCF — `calc_svcf()`

`calc_svcf()` applies the copy-number–aware correction and produces the final SVCF estimates.

The sign of the SV-first candidate supplies the mathematical ordering criterion. The `thresh` parameter (default `0.1`) adds an operational noise buffer around zero: when `ss1 <= thresh`, SVCFit uses the alternate candidate. Deletions use the alternate candidate unconditionally. The alternate estimate is constrained to 0–1 after the zygosity correction, and the unconstrained value and boundary status remain in the output.

```{r}
#| label: calcsvcf
svcfit_out2 <- calc_svcf(anno_sv_cnv, sv_info, thresh = 0.1,
                          samp = "example_sample", exper = "exp1")
head(svcfit_out2 %>% select(ID, classification, raw_svcf, final_svcf))
```

------------------------------------------------------------------------

# Hemizygous chromosomes (chrX / chrY)

Autosomal SVCF estimation relies on heterozygous germline SNPs to infer the allele copy ratio and phase each SV against overlapping CNVs. A **hemizygous** chromosome — a male X or Y, or any chromosome single-copy in the germline — carries no heterozygous SNPs, so that route is undefined and the diploid conversion factor of 2 does not apply. Uncorrected, a clonal hemizygous SV is estimated at roughly twice its true cellular fraction and then drops out of the clustered output.

SVCFit corrects this with a ploidy-aware estimator keyed on `cn_bar`, the mean copy number of the locus per cell, measured from read depth. With `VAF = BPC / (BPC + BEC)`:

-   **SV precedes CNV (H1):** `SVCF = cn_bar * VAF - (cn_bar - 1)`
-   **CNV precedes SV (H2):** `SVCF = cn_bar * VAF`

Both reduce to `SVCF = VAF` when `cn_bar = 1` (copy-neutral). The ordering is chosen by the sign of the H1 form. A tandem duplication uses `SVCF = (cn_bar - 1) / (r - 1)` and is reported as an upper bound, because `r` is not identifiable from a single locus.

## Running the pipeline on a hemizygous chromosome

Pass the single-copy chromosomes and a per-SV `cn_bar` table (from a read-depth segmentation) to `run_svcfit()`:

```{r}
#| label: hemizygous-run
#| eval: false
result <- run_svcfit(
  p_het = p_het, p_onsv = p_onsv, p_sv = p_sv, p_cnv = p_cnv,
  samp = "SampleID", exper = "ExperimentID",
  hemizygous_chr = c("chrX", "chrY"),   # single-copy in this subject's germline
  hemi_cn_bar    = cn_bar_table         # data.frame(CHROM, POS, cn_bar)
)
```

With `hemizygous_chr = NULL` the results are identical to the diploid path. With `hemi_cn_bar = NULL`, hemizygous SVs are treated as copy-neutral and resolve to `SVCF = VAF` — except tandem duplications, which need `cn_bar` and are left unresolved (`final_svcf = NA`, flagged `hemizygous_dup_needs_cn_bar` in `svcf_status`). A deletion or inversion on a copy-altered segment cannot be singled out without `cn_bar` on a hemizygous chromosome, so it too falls back to `SVCF = VAF`.

## Where `cn_bar` comes from (DNAcopy depth segmentation)

SVCFit consumes `cn_bar` but does not compute it. FACETS — used for autosomal copy number — cannot fit a hemizygous chromosome: with no heterozygous germline SNPs it returns a single whole-chromosome segment with an inflated, unusable total copy number. `cn_bar` is instead measured from **read depth** and segmented with the Bioconductor **DNAcopy** package (circular binary segmentation):

1.  Bin chrX read depth (e.g. 100 kb) in the tumor and the matched-normal BAM.
2.  Normalize each BAM by its own autosomal 2-copy baseline and take the tumor/normal ratio; because the matched normal is also single-copy on chrX, this ratio is the tumor's mean copy number per cell.
3.  Scale by `psi_sample / 2` (`psi_sample = purity * psi_tumor + (1 - purity) * 2`, from the FACETS autosomal fit) to recover absolute copy number.
4.  Segment `log2(cn_bar)` with DNAcopy; each segment's `cn_bar = 2^seg.mean`.

```{r}
#| label: hemizygous-cnbar
#| eval: false
library(DNAcopy)
# cn_bar_bin : per-bin (tumor/normal depth ratio) * (psi_sample / 2)
# pos        : bin start positions on the hemizygous contig
cna <- CNA(log2(cn_bar_bin), rep("chrX", length(pos)), pos,
           data.type = "logratio", sampleid = "chrX")
seg <- segment(smooth.CNA(cna), alpha = 0.01, min.width = 2)$output
seg_cn_bar <- 2^seg$seg.mean   # mean copy number per cell, per segment
```

`cn_bar` is never rounded to an integer — the hemizygous forms take it directly.

**DNAcopy is replaceable.** SVCFit never sees DNAcopy or its segment objects; it consumes a plain per-SV table. Any source of a mean copy number per cell — another segmenter (PSCBS, HMMcopy, GATK ModelSegments, CNVkit, Battenberg/ASCAT) or even a single tumor/normal depth ratio per breakpoint — works once coerced into `hemi_cn_bar`: a `data.frame` with columns `CHROM` (contig, matching the SV exactly), `POS` (the SV breakpoint, matching **exactly**), and `cn_bar` (mean copies per cell). The lookup is an exact `(CHROM, POS)` match — **one row per SV, keyed on the breakpoint**, not per segment. If your source gives segment *intervals*, assign each SV the `cn_bar` of the segment spanning its `POS` first. An SV with no matching row gets no depth correction — a tandem duplication is left unresolved, everything else falls back to `SVCF = VAF`. A bare numeric vector is also accepted, but it is recycled across all SV rows (not just hemizygous ones), so only a single value — the same `cn_bar` for every hemizygous SV — is reliable. The optional `hemi_bg_cn` uses the same convention with columns `CHROM, POS, bg_cn`.

```{r}
#| label: hemizygous-cnbar-format
#| eval: false
# however you produced cn_bar, this per-SV table is all SVCFit reads
hemi_cn_bar <- data.frame(
  CHROM  = c("chrX",    "chrX"),
  POS    = c(31200000L, 67500000L),   # exact SV breakpoint positions
  cn_bar = c(1.02,      1.54)         # mean copies/cell at each locus
)
```

## Estimator helpers

The estimator building blocks are exported, so read counts and a `cn_bar` can be scored directly without the VCF pipeline:

```{r}
#| label: hemizygous-helpers
# local normal ploidy: 1 on a hemizygous chromosome, 2 on autosomes
local_ploidy(c("chr1", "chrX", "chrY"), hemizygous_chr = c("chrX", "chrY"))

# an SV inside a CNV made by another event — ordering picked by the sign rule
resolve_hemizygous_svcf(bpc = 4, bec = 1, cn_bar = 2.0)

# a copy-neutral locus reduces to SVCF = VAF
resolve_hemizygous_svcf(bpc = 3, bec = 7, cn_bar = 1.0)

# a hemizygous tandem duplication (reported as an upper bound)
hemizygous_dup_svcf(cn_bar = 1.5, r = 2)
```

------------------------------------------------------------------------

# Individual sub-functions

Both `extract_info()` and `characterize_sv()` are wrappers. You can call their constituent steps individually for debugging or custom pipelines.

## `extract_info()` sub-functions

### `load_data()`

Reads all four input files and returns a named list.

```{r}
#| label: load
data <- load_data(p_het, p_onsv, p_sv, p_cnv, chr = NULL, tumor_only = FALSE)
# Returns: data$het_snp, data$het_on_sv, data$sv, data$cnv
```

### `proc_bnd()`

Classifies BND (breakend) records into three translocation types:

| Type | Description |
|-----------------------|-------------------------------------------------|
| `copbnd` | Copy-paste (unbalanced): one copy is inserted elsewhere, original is retained |
| `cutbnd` | Cut-paste (unbalanced): original locus is deleted |
| `rtl` | Reciprocal translocation: balanced exchange between two loci |

```{r}
#| label: procbnd
bnd_info <- proc_bnd(sv = data$sv, flank_del = 50)
```

### `parse_sv_info()`

Extracts SV type, genomic coordinates, and read counts; applies QUAL and read-count filters.

```{r}
#| label: parsesv
sv_info <- parse_sv_info(sv = data$sv, bnd = bnd_info$bnd, del = bnd_info$del,
                          QUAL_thresh = 100, min_alt = 2)
```

### `parse_het_snps()`

Extracts allele counts from the heterozygous SNP VCF.

```{r}
#| label: parsesnps
snp_df <- parse_het_snps(data$het_snp)
```

### `parse_snp_on_sv()`

Compares SNP allele counts in all reads vs. SV-supporting reads to determine which haplotype the SV is on.

```{r}
#| label: parseonsv
sv_phase <- parse_snp_on_sv(het_on_sv = data$het_on_sv, snp_df)
```

## `characterize_sv()` sub-functions

### `assign_svids()`

Assigns each heterozygous SNP to the nearest SV within `flank_snp` bp.

```{r}
#| label: assignids
assign_id <- assign_svids(sv_phase, sv_info, flank_snp = 500)
```

### `sum_sv_info()`

Summarises the per-SNP phase evidence into a single zygosity and phase call per SV.

```{r}
#| label: sumsv
sv_sum <- sum_sv_info(sv_phase, assign_id, sv_info)
```

### `assign_cnv()`

Overlaps SV breakpoints with CNV segments to find which segment (if any) each SV falls in.

```{r}
#| label: assigncnv
sv_cnv <- assign_cnv(sv_sum, data$cnv)
```

### `annotate_cnv()`

Classifies each overlapping CNV as a deletion or duplication using copy-number values.

```{r}
#| label: annotcnv
anno_sv_cnv <- annotate_cnv(sv_cnv)
```

------------------------------------------------------------------------

# Downstream: clustering and tree inference — `build_trees()`

After computing SVCF values across multiple samples (e.g. pre-treatment and on-treatment biopsies), you can cluster SVs into subclones and infer a phylogenetic tree.

`build_trees()` runs in two stages:

1.  **DP-GMM clustering** (`run_clustering = TRUE`): groups SVs with similar SVCF profiles across paired samples into clones using a Dirichlet-process Gaussian mixture model. Requires Python with `scikit-learn` installed (via the `reticulate` package).
2.  **Tree inference** (`run_tree = TRUE`): fits a spanning tree over the clone CCF table that satisfies the lineage-precedence constraint (parent CCF ≥ sum of children CCFs).

```{r}
#| label: build-trees-example
#| eval: false
# Example — requires paired BED files and a purity table
bt <- build_trees(
  run_clustering = TRUE,
  pair_path      = "pairs.txt",   # tab-delimited: pre_BAT, on_BAT
  pur_path       = "purity.txt",  # tab-delimited: sample, purity
  data_dir       = "/path/to/svcfit_output",
  pair_num       = 1L,
  run_tree       = TRUE
)

# bt[[1]] — clustering results
# bt[[2]][[3]] — igraph plot of the inferred phylogenetic tree
bt[[2]][[3]]
```

------------------------------------------------------------------------

# Optional: evaluating accuracy on simulated data

`load_truth()` and `attach_truth()` are only applicable when the ground-truth clone assignments are known (e.g. VISOR simulations). They attach the true clone label to each SV so you can measure how accurately SVCFit recovered the cellular fraction.

```{r}
#| label: truth
#| message: false
overlap      <- FALSE   # TRUE only for overlapping-clone simulations
truth        <- load_truth(p_truth, overlap)
svcfit_truth <- attach_truth(svcfit_out, truth)

# Show estimated SVCF alongside the ground-truth clone label
# clone: "clonal" = present in all simulated cancer cells,
#        "sub"    = present in one subclone only
svcfit_truth %>%
  select(ID, classification, final_svcf, clone) %>%
  head(6)
```

------------------------------------------------------------------------

# Session info

```{r}
#| label: session
sessionInfo()
```
